### 2013-01-14 sections/evaluation.tex
changed NFQL tool to \texttt{nfql}

It can be seen that the performance of the filter stage in \ac{NFQL} is
comparable to that of \texttt{flowtools} and SiLK. SiLK takes less time on
lower ratios, but then SilK and \texttt{nfdump} also use their own file
format. \texttt{nfdump} appears to be significantly faster than the rest. This
is because \texttt{nfdump} uses the \texttt{lzo} compression scheme. Adding
\texttt{lzo} compression will likely improve \ac{NFQL}'s filter
performance. Note that all the tools were single-threaded
in this evaluation, and did not completely utilize the $24$ available cores.

@@

The evaluation graph reveals that the time taken by the tools are comparable
on lower ratios, but on higher ratios, \ac{NFQL} starts to drift apart. Since
most of the time is taken in writing the records to files, it is unclear
whether SiLK's usage of its own file format is responsible for the drift.
SiLK's \texttt{rwgroup} tool is also supplied a \texttt{-{}-summarize} flag to
force it to write only the first record of each group, to make both tools
write the same number of records. This gives SiLK the leverage to not store
information about which members are part of the group. \ac{NFQL} on the other
hand needs to allocate resources (which may take time) to keep this
information in its data structures, since the ungrouper later may request to
write the members of a group while unfolding the tuples.  It is also important
to note that both the tools again remained single-threaded throughout the
evaluation. SiLK took advantage of an inherent concurency arising from a pipe
between \texttt{rwsort} and \texttt{rwgroup}, which makes the two processes
run concurrently, the effect of which gets more pronounced on higher ratios.
The profiling results from GNU \texttt{gprof} \cite{graham:1982} indicate that
$60\%$ of the time is taken in \texttt{qsort} comparator calls.  As a result,
it comes as no surprise, that bifurcating \texttt{qsort} invocation to
multiple threads and later merging the results back using merge sort will help
parallelize the grouper stage and maybe reduce the drift on higher ratios. In
addition, since all of the evaluation queries had grouping terms using an
equality comparator, \ac{NFQL} can introspect such a grouping rule to
dynamically optimize processing searches using a hashtable and turn to
\texttt{qsort} based grouping only as a fallback.

@@

It can be seen that the timings of \ac{NFQL} are far apart from that of SiLK.
It is due to the drift already created by the grouper at the $1.0$ ratio in
the previous stage. As a result, the group filter comes into play only after
$300$ seconds, whereas SiLK's group filtering already starts just below $150$
seconds. Even if we normalize the graph, it can be observed that the \ac{NFQL}
group filter has a higher slope. This is because it is only executed once the
grouper returns, and therefore has to reiterate the groups to make a filtering
decision.

@@

It can be seen that the merger is the most performance critical stage of the
\ac{NFQL} pipeline thus far. It is due to the fact that the merger is working
on twice the number of flow records than any other previous stage. In
addition, each branch is writing the results of the filter, grouper and group
filter stage to \texttt{flowtools} files. As a result, the amount of disk I/O
involved is twice as much as well. Even though each branch is delegated to a
separate core using affinity masks, most of time is taken in writing these
results to the file. These results although look less promising, they are way
better than the original \ac{NFQL} merger implementation. The optimized merger
implementation takes advantage of sorted nature of filtered groups and
therefore can significantly reduce the number of merger matches. It also
writes a merged group record to a file only once despite the number of times
it has matched. Without these optimizations, running such queries on the
merger kept the CPU churning for days without results.

>>>>

It can be seen that the performance of the filter stage in \texttt{nfql} is
comparable to that of \texttt{flowtools} and SiLK. SiLK takes less time on
lower ratios, but then SilK and \texttt{nfdump} also use their own file
format. \texttt{nfdump} appears to be significantly faster than the rest. This
is because \texttt{nfdump} uses the \texttt{lzo} compression scheme. Adding
\texttt{lzo} compression will likely improve \texttt{nfql}'s filter
performance. Note that all the tools were single-threaded
in this evaluation, and did not completely utilize the $24$ available cores.

@@

The evaluation graph reveals that the time taken by the tools are comparable
on lower ratios, but on higher ratios, \texttt{nfql} starts to drift apart.
Since most of the time is taken in writing the records to files, it is unclear
whether SiLK's usage of its own file format is responsible for the drift.
SiLK's \texttt{rwgroup} tool is also supplied a \texttt{-{}-summarize} flag to
force it to write only the first record of each group, to make both tools
write the same number of records. This gives SiLK the leverage to not store
information about which members are part of the group. \texttt{nfql} on the
other hand needs to allocate resources (which may take time) to keep this
information in its data structures, since the ungrouper later may request to
write the members of a group while unfolding the tuples.  It is also important
to note that both the tools again remained single-threaded throughout the
evaluation. SiLK took advantage of an inherent concurency arising from a pipe
between \texttt{rwsort} and \texttt{rwgroup}, which makes the two processes
run concurrently, the effect of which gets more pronounced on higher ratios.
The profiling results from GNU \texttt{gprof} \cite{graham:1982} indicate that
$60\%$ of the time is taken in \texttt{qsort} comparator calls.  As a result,
it comes as no surprise, that bifurcating \texttt{qsort} invocation to
multiple threads and later merging the results back using merge sort will help
parallelize the grouper stage and maybe reduce the drift on higher ratios. In
addition, since all of the evaluation queries had grouping terms using an
equality comparator, \texttt{nfql} can introspect such a grouping rule to
dynamically optimize processing searches using a hashtable and turn to
\texttt{qsort} based grouping only as a fallback.

@@

It can be seen that the timings of \texttt{nfql} are far apart from that of
SiLK.  It is due to the drift already created by the grouper at the $1.0$
ratio in the previous stage. As a result, the group filter comes into play
only after $300$ seconds, whereas SiLK's group filtering already starts just
below $150$ seconds. Even if we normalize the graph, it can be observed that
the \texttt{nfql} group filter has a higher slope. This is because it is only
executed once the grouper returns, and therefore has to reiterate the groups
to make a filtering decision.

@@

It can be seen that the merger is the most performance critical stage of the
\ac{NFQL} pipeline thus far. It is due to the fact that the merger is working
on twice the number of flow records than any other previous stage. In
addition, each branch is writing the results of the filter, grouper and group
filter stage to \texttt{flowtools} files. As a result, the amount of disk I/O
involved is twice as much as well. Even though each branch is delegated to a
separate core using affinity masks, most of time is taken in writing these
results to the file. These results although look less promising, they are way
better than the original \texttt{nfql} merger implementation. The optimized
merger implementation takes advantage of sorted nature of filtered groups and
therefore can significantly reduce the number of merger matches. It also
writes a merged group record to a file only once despite the number of times
it has matched. Without these optimizations, running such queries on the
merger kept the CPU churning for days without results.

### 2013-01-14 sections/implementation.tex
changed NFQL tool to \texttt{nfql}

The \ac{NFQL} architecture primarily consists of a front-end parser backed up
by an execution engine as shown in Fig. \ref{fig:nfql-pipeline}. The execution
engine is the brain of \ac{NFQL} where the complete pipeline is processed. It
receives the flow query at runtime using a JSON \cite{rfc4627} intermediate
format. The execution engine is written in C, however the front-end parser can
be written in any desired language.

%\begin{figure}[ht!]
  %\begin{center}
    %\includegraphics* [width=0.8\linewidth]{nfql-architecture}
    %\caption{NFQL Architecture}
    %\label{fig:architecture}
  %\end{center}
%\end{figure}

>>>>

The \texttt{nfql} architecture primarily consists of a front-end parser backed
up by an execution engine as shown in Fig. \ref{fig:nfql-pipeline}. The
execution engine is the brain of \texttt{nfql} where the complete pipeline is
processed. It receives the flow query at runtime using a JSON \cite{rfc4627}
intermediate format. The execution engine is written in C, however the
front-end parser can be written in any desired language.

### 2013-01-14 sections/implementation/execution-workflow.tex
changed NFQL tool to \texttt{nfql}

\subsubsection{Splitter} \ac{NFQL} uses identifiers to reference a flow record
in the \texttt{char} array. This eliminated the need to copy all the
flow-records when moving ahead in the pipeline. As a result, there is no
dedicated splitter stage in the execution engine. Each branch references the
flow records from a common memory location. This helps keep the memory costs
at a minimum when multiple branches are involved.

\subsubsection{Filter}
\begin{figure}[h!]
  \begin{center}
    \includegraphics* [width=0.9\linewidth]{filter-fv1-fv2}
    \caption{Filter Stage: Flowy 2.0 vs NFQL}
    \label{fig:fv1-fv2-filter}
  \end{center}
\end{figure}

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to produce the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \ac{NFQL} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a flow-tools file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. We used the publicly available Trace 7 from the
SimpleWeb \cite{simpleweb} to compare the performance of \ac{NFQL} against the
one defined by the flow query language as shown in Fig.
\ref{fig:fv1-fv2-filter}. The filter stage implementation with these
optimizations runs 10 times faster and is more pronounced on higher ratios.

\begin{figure}[ht!]
  \begin{center}
    \includegraphics*[width=0.9\linewidth]{grouper-fv2-bsearch-fv2}
    \caption{Grouper Stage: NFQL (Generic) vs NFQL (Specific)}
    \label{fig:fv1-fv2-grouper}
  \end{center}
\end{figure}

@@

In order to be able to make relative comparisons on field offsets, a simple
approach is to linearly walk through each filtered record against the filtered
recordset leading to a complexity of $O(n^2)$, where $n$ is the number of
filtered records. A smarter approach is to put the copy in a hash table and
then try to map each pointer while walking down the filtered recordset once,
leading to a complexity of $O(n)$. The hash table approach however, will only
work on equality comparisons. A better approach, as implemented in \ac{NFQL}
is to sort the filtered recordset on the field offsets of all the requested
grouping terms of a clause. This helps the execution engine perform a nested
binary search to reduce the linear pass to a fairly small filtered recordset.
This helps the grouper perform faster search lookups to find records that must
group together in $O(n*lg(k))$ time with a preprocessing step taking
$O(p*n*lg(n))$ in the average case, where $n$ is the number of filtered
records, $p$ is the number of grouping terms in a clause and $k$ is the number
of unique filtered records. The grouping approach has further been optimized
when the filtered records are grouped for equality. In such a scenario, the
need to search for unique records and a  subsequent binary search goes away.
The groups can now be formed in $O(n)$ time with the same preprocessing step
taking $O(p*n*lg(n))$.  The performance evaluation of the grouper handling
this special case against when handling generic cases is shown in Fig.
\ref{fig:fv1-fv2-grouper}.

@@

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records, however, can be reduced by sorting the group records on the
field offsets used for a match. \ac{NFQL} optimizes the merger to skip over
iterator permutations when a state of a current field offset value may not
allow any further match beyond the index in the current branch.  For such an
optimization to work, the filtered group records must be sorted in the order
of field offsets specified in the merger clause.  Specifying the filtered
group records in any other order may lead to undefined behavior. This means,
that if the same field offsets were used in the grouper stage, the terms in
the group clause can be rearranged by the query designer to align with the
order of terms in the merger clause.

>>>>

\subsubsection{Splitter} \texttt{nfql} uses identifiers to reference a flow
record in the \texttt{char} array. This eliminated the need to copy all the
flow-records when moving ahead in the pipeline. As a result, there is no
dedicated splitter stage in the execution engine. Each branch references the
flow records from a common memory location. This helps keep the memory costs
at a minimum when multiple branches are involved.

\subsubsection{Filter}
\begin{figure}[h!]
  \begin{center}
    \includegraphics* [width=0.9\linewidth]{filter-fv1-fv2}
    \caption{Filter Stage: Flowy 2.0 vs \texttt{nfql}}
    \label{fig:fv1-fv2-filter}
  \end{center}
\end{figure}

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to produce the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \texttt{nfql} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a flow-tools file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. We used the publicly available Trace 7 from the
SimpleWeb \cite{simpleweb} to compare the performance of \texttt{nfql} against
the one defined by the flow query language as shown in Fig.
\ref{fig:fv1-fv2-filter}. The filter stage implementation with these
optimizations runs 10 times faster and is more pronounced on higher ratios.

\begin{figure}[ht!] \begin{center}
    \includegraphics*[width=0.9\linewidth]{grouper-fv2-bsearch-fv2}
    \caption{Grouper Stage: \texttt{nfql} (Generic) vs \texttt{nfql}
  (Specific)} \label{fig:fv1-fv2-grouper} \end{center} \end{figure}

@@

In order to be able to make relative comparisons on field offsets, a simple
approach is to linearly walk through each filtered record against the filtered
recordset leading to a complexity of $O(n^2)$, where $n$ is the number of
filtered records. A smarter approach is to put the copy in a hash table and
then try to map each pointer while walking down the filtered recordset once,
leading to a complexity of $O(n)$. The hash table approach however, will only
work on equality comparisons. A better approach, as implemented in
\texttt{nfql} is to sort the filtered recordset on the field offsets of all
the requested grouping terms of a clause. This helps the execution engine
perform a nested binary search to reduce the linear pass to a fairly small
filtered recordset.  This helps the grouper perform faster search lookups to
find records that must group together in $O(n*lg(k))$ time with a
preprocessing step taking $O(p*n*lg(n))$ in the average case, where $n$ is the
number of filtered records, $p$ is the number of grouping terms in a clause
and $k$ is the number of unique filtered records. The grouping approach has
further been optimized when the filtered records are grouped for equality. In
such a scenario, the need to search for unique records and a  subsequent
binary search goes away.  The groups can now be formed in $O(n)$ time with the
same preprocessing step taking $O(p*n*lg(n))$.  The performance evaluation of
the grouper handling this special case against when handling generic cases is
shown in Fig.  \ref{fig:fv1-fv2-grouper}.

@@

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records, however, can be reduced by sorting the group records on the
field offsets used for a match. \texttt{nfql} optimizes the merger to skip
over iterator permutations when a state of a current field offset value may
not allow any further match beyond the index in the current branch.  For such
an optimization to work, the filtered group records must be sorted in the
order of field offsets specified in the merger clause.  Specifying the
filtered group records in any other order may lead to undefined behavior. This
means, that if the same field offsets were used in the grouper stage, the
terms in the group clause can be rearranged by the query designer to align
with the order of terms in the merger clause.

### 2013-01-14 sections/implementation/performance-optimizations.tex
changed NFQL tool to \texttt{nfql}

There can be a situation where the query designer may incorrectly ask for
aggregation on a field already specified in a grouper (or filter) clause. If
the relative operator is an equality comparison, the aggregation on such a
field becomes less useful, since the members of the grouped record will always
have the same value for that field. \ac{NFQL} realizes this redundant request
and ignores such aggregations.

\ac{NFQL} has dedicated comparator functions for each type of operation and
the type of the field offset it operates upon. It is not guaranteed that given
the type of the query and trace, the program will eventually go through each
stage of the pipeline. It is also possible that the program exits before,
because there is nothing more for the next stage to compute. The function
pointers are therefore set as late as possible and are called from their
respective stages just before the comparison is needed.

>>>>

There can be a situation where the query designer may incorrectly ask for
aggregation on a field already specified in a grouper (or filter) clause. If
the relative operator is an equality comparison, the aggregation on such a
field becomes less useful, since the members of the grouped record will always
have the same value for that field. \texttt{nfql} realizes this redundant
request and ignores such aggregations.

\texttt{nfql} has dedicated comparator functions for each type of operation
and the type of the field offset it operates upon. It is not guaranteed that
given the type of the query and trace, the program will eventually go through
each stage of the pipeline. It is also possible that the program exits before,
because there is nothing more for the next stage to compute. The function
pointers are therefore set as late as possible and are called from their
respective stages just before the comparison is needed.

### 2013-01-14 sections/relatedwork.tex
back to 6 pages
changed NFQL tool to \texttt{nfql}

SiLK \cite{SiLK} is a network traffic collection and analysis tool that comes
quite close to providing similar capabilities as \ac{NFQL} and is therefore
used as a reference point to compare the performance of the \ac{NFQL}
execution engine. The design and implementation of SiLK, however, differs a
lot from that of \ac{NFQL}. For instance, in SiLK there are separate tools to
perform the task of each stage of the \ac{NFQL} processing pipeline. The stage
functionality is not full-fledged though. The grouping and merging operations
can only be performed using an equality operator. This restriction allows the
tool to perform optimization such as using hash tables to perform lookups.
\ac{NFQL} on the other hand, provides a much richer set of comparisons
operations, such as \emph{equal, not equal, greater than, less than, greater
or equal, less or equal}. There are also stringent requirements to how
flow-data needs to be organized before it can be piped into a tool. The
grouping tool, for instance, assumes that the to-be supplied input flow data
is already sorted on the field column.  These requirements can make it a
little cumbersome to design a full-fledged flow query.  For instance, trying
to mimic a \ac{NFQL} query in SiLK sometimes ends up as a shell script with
over a dozen of SiLK tools piped together.

>>>>

SiLK \cite{SiLK} is a network traffic collection and analysis tool that comes
quite close to providing similar capabilities as \ac{NFQL} and is therefore
used as a reference point to compare the performance of the \texttt{nfql}
execution engine. The design and implementation of SiLK, however, differs a
lot from that of \texttt{nfql}. For instance, in SiLK there are separate tools to
perform the task of each stage of the \ac{NFQL} processing pipeline. The stage
functionality is not full-fledged though. The grouping and merging operations
can only be performed using an equality operator. This restriction allows the
tool to perform optimization such as using hash tables to perform lookups.
\ac{NFQL} on the other hand, provides a much richer set of comparisons
operations, such as \emph{equal, not equal, greater than, less than, greater
or equal, less or equal}. There are also stringent requirements in SiLK how
flow-data needs to be organized before it can be piped into a tool. The
grouping tool, for instance, assumes that the to-be supplied input flow data
is already sorted on the field column.  These requirements can make it a
little cumbersome to design a full-fledged flow query.  For instance, trying
to mimic a \ac{NFQL} query in SiLK sometimes ends up as a shell script with
over a dozen of SiLK tools piped together.
